---
title: "Final Project - Predicting Annual Air Pollution"
author: "Xuan Jiang, Yutong Guo, Tengyue Wang"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 72
---

## Introduction

This project aims to harness the power of data analytics, utilizing R,
to forecast annual average air pollution concentrations across the
United States at the granularity of zip code regions. Air pollution
remains a critical environmental challenge impacting public health,
ecosystem stability, and global climate change. Its complex nature,
influenced by a myriad of factors ranging from urbanization to chemical
atmospheric interactions, necessitates advanced predictive models for
effective management and policy planning. Air pollution, comprising
various pollutants like fine particles (PM2.5), ozone (O3), nitrogen
oxides (NOx), and black carbon, has been associated with a range of
adverse health outcomes, including asthma, inflammation in Type 1
diabetes, lung diseases, low birth weight, and viral Infection.[1]

Traditional epidemiological approaches to assessing air pollution levels
have relied on a network of stationary monitors, which are often
insufficient in number and unevenly distributed, particularly in rural
or underdeveloped areas. Moreover, these monitors are not strategically
placed near known pollution sources, leading to potential
underestimations of exposure levels, which is known as sparse
monitoring. These limitations hinder the ability to accurately assess
and understand the full impact of air pollution on public health.2
However, machine learning methods offer a promising solution for
predicting air pollution levels. These advanced computational techniques
can analyze large datasets, identifying complex patterns that are not
evident through traditional analysis. By leveraging various types of
data sources, such as satellite imagery, traffic patterns, industrial
activity data, and meteorological information, machine learning
algorithms can generate accurate predictions of air pollution levels in
specific areas, even in the absence of direct monitoring data, which is
particularly valuable in filling the gaps left by sparse monitoring
networks.[2]

### Load packages

```{r load-packages, message=FALSE}
library(tidyverse)
library(tidymodels)
library(maps)
library(lmtest)
library(randomForest)
#install.packages("maps")
#install.packages("OCSdata")
#install.packages("here")
#install.packages("skimr")
```

## Research Question

1.  Can we predict US annual average air pollution concentrations at the
    granularity of zip code regional levels using predictors such as
    data about population density, urbanization, and road density, as
    well as, satellite pollution data and chemical modeling data?

2.  How do varying levels of air pollutant concentrations across
    different states correlate with the mortality rates from respiratory
    conditions within those states？

## The Data

The dataset used in this study is an integration of data from "US
Environmental Protection Agency (EPA), the National Aeronautics and
Space Administration (NASA), the US Census, and the National Center for
Health Statistics (NCHS)," which was initially collected by a researcher
group at the Johns Hopkins School of Public Health.[3] This dataset,
originating from 876 air quality monitors, provides a comprehensive view
of PM2.5 (particulate matter with a diameter of less than 2.5
micrometers) concentrations across various locations. It consists of 50
columns, encompassing a wide range of environmental, geographical, and
socio-demographic variables.

The dataset we used for extended data analysis is "Weekly Counts of
Deaths by State and Select Causes, 2014-2019", which was compiled by the
National Center for Health Statistics (NCHS) and published on April 1st,
2022. It records the number of deaths, categorized by specific causes,
on a weekly basis. We focus on the column that contains the death counts
of respiratory diseases in this study.[4]

### Data Import

```{r}
# Load the dataset 
# OCSdata::raw_data("ocs-bp-air-pollution", outpath = getwd())
df <- readr::read_csv(here::here("OCS_data", "data","raw", "pm25_data.csv"))
# View the dataset 
df
```

### Data Wrangling

```{r}
# Get an overall sense of the dataset
skimr::skim(df)
# Select the columns that we needed and rename the dataset to "pm"
pm <- subset(df, select = c(value, state, county, lat, lon, county_pop, log_prisec_length_1000,	popdens_county, popdens_zcta, pov, urc2013, aod, CMAQ ))
# Rename the column name to aid understanding 
pm <- pm |> rename("log_RoadLength_1kmRadius" = "log_prisec_length_1000",
                   "optical_depth" = "aod")
# View the wrangled dataset "pm"
view(pm)
```

### Exploratory Data Analysis

In approaching this question, we first visualize the distribution of
PM2.5 values across states to gain a broad overview of air quality
throughout the United States. This approach can highlight regional
variations and potential hotspots of pollution. By mapping PM2.5 levels
on a state-by-state basis, we can swiftly identify areas with higher
pollution levels. This provides valuable context before we delve into a
more granular analysis at the zip code level.

**Figure1:** Spatial Distribution of PM 2.5 Values

```{r}
ggplot(pm, aes(x = lon, y = lat, color = value)) +
  borders("state", colour = "black", fill = "white") + 
  geom_point(alpha = 0.5) +
  scale_color_viridis_c() +
  labs(title = "Spatial Distribution of PM 2.5 Values",
       x = "Longitude", y = "Latitude", color = "PM 2.5 Value") +
  theme_minimal()
```

**Result:** Based on the longitude and latitude of the distribution of
the united states,We can see the California is extremely polluted, there
are yellow dots on both North and South ends of California. Cities in
the middle is a lot less polluted, probably because it is more rural
compared to the coasts.

**Figure2:** Spatial Distribution of CMAQ Values

```{r}
ggplot(pm, aes(x = lon, y = lat, color = CMAQ)) +
  borders("state", colour = "black", fill = "white") + 
  geom_point(alpha = 0.5) +
  scale_color_viridis_c() +
  labs(title = "Spatial Distribution of CMAQ Values",
       x = "Longitude", y = "Latitude", color = "CMAQ Values") +
  theme_minimal()
```

**Result:** The results shown in Figure 2 are almost identical to those
in Figure 1, indicating that the air pollution data generated by the
'Community Multiscale Air Quality' (CMAQ) computational model accurately
represents PM2.5 pollution. This similarity suggests a high level of
accuracy and reliability in the model's ability to simulate air
pollution scenarios, making the CMAQ value a potentially strong
predictor.

**Figure3:** The box-plot of PM 2.5 Values of the state

```{r}
ggplot(pm, aes(x = state, y = value)) +
  geom_boxplot() +
  labs(title = "PM2.5 Values by State",
       x = "State",
       y = "PM2.5 Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Then we created a boxplot to compare the pm 2.5 value of each state.
This way we have a better idea of the distribution of the data, which
would serve well in the later extended analysis while comparing state
air pollution level by death.

**Result:** The boxplot provides a comparative view of PM2.5 values
across several states. States like California, Pennsylvania, and New
York show a wider range of PM2.5 concentrations, as indicated by their
longer box and whisker spans, which points to a higher variability in
air pollution within these states. Outliers are particularly noticeable
in states such as Utah, suggesting sporadic instances of very high PM2.5
levels. On the other hand, states like Vermont and New Hampshire display
shorter boxes and whiskers, reflecting a more consistent PM2.5 level
with less variability. The median PM2.5 levels, marked by the horizontal
line within the boxes, vary from state to state, with some states
displaying medians on the higher end of the scale, potentially signaling
an overall higher level of air pollution.

**Figure4:** Relationship between PM 2.5 values and poverty 
```{r}
#PM 2. 5 levels versus poverty 
ggplot(df, aes(x = pov, y = value)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Relationship between PM 2.5 Levels and Poverty Rate",
       x = "Poverty Rate (%)", y = "PM 2.5 Value") +
  theme_minimal()
#PM 2.5 levels versus popdens_zcta
ggplot(df, aes(x = popdens_zcta, y = value)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Relationship between PM 2.5 Levels and popdens_zcta",
       x = "Popdens_zcta", y = "PM 2.5 Value") +
  theme_minimal()
#PM 2.5 levels versus Aerosol Optical Depth measurement from a NASA satellite
ggplot(df, aes(x = aod, y = value)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Relationship between PM 2.5 Levels and Aod",
       x = "Aerosol Optical Depth measurement from a NASA satellite", y = "PM 2.5 Value") +
  theme_minimal()
```
**Result:** 
**Poverty Rate**:
The scatter plot indicates a positive correlation between PM 2.5 levels, a measure of air pollution, and poverty rates. As poverty rates increase, there appears to be a corresponding rise in PM 2.5 levels, suggesting that areas with higher poverty may also suffer from higher air pollution. 
**Population density (number of people per kilometer squared area of zcta)**:
The provided scatter plot depicts the relationship between PM 2.5 levels and population density (popdens_zcta). The plot shows a general upward trend, as indicated by the blue trend line, suggesting that higher population densities tend to correspond with higher PM 2.5 levels. This could imply that more densely populated areas experience greater air pollution. The shaded area around the trend line may represent a confidence interval, indicating where the true relationship is likely to lie with a certain level of certainty. 
**Aerosol Optical Depth measurement from a NASA satellite**:
The scatter plot illustrates a positive relationship between PM 2.5 leveles and aerosol optical depth (AOD), the blue line, likely representing a linear regression fit, shows that as AOD increases, PM 2.5 levels tend to increase as well. This suggests that higher concentrations of aerosols in the atmosphere are associated with higher concentrations of PM 2.5. The shaded area around the line may indicate the confidence interval, providing a range of values where the true line of best fit is expected to fall. The data points are somewhat tightly clustered along the trend line, especially at lower AOD values, which indicates a stronger relationship between these two variables in that range. 

## Data Analysis

```{r}
#correlation matrix
cor_matrix <- cor(pm |>
                    select(popdens_zcta, urc2013, log_RoadLength_1kmRadius, optical_depth, CMAQ, value))
cor_matrix
```

We can first run a correlation matrix between the 5 main factors and the
pm 2.5 value to see wheather there is a significant correlation between
each factor and the pm 2.5 value.

**Result**: Looking at the last column of this table: The slight
positive correlation between population density and PM2.5 values
(0.1330540) suggests that areas with higher numbers of people could have
increased emissions from sources like vehicles and industrial
activities, contributing to higher PM2.5 levels. Despite being a weak
correlation, population density still carries potential predictive value
as a proxy for human-related pollution sources.The modest negative
correlation between urbanization level and PM2.5 values (-0.2508474)
indicates an inverse relationship where more urbanized areas could
exhibit lower PM2.5 levels. This could be attributed to better
infrastructure for pollution control and regulation enforcement in
highly urbanized areas. Although the result is counter-intuitive, the
variable urbanization level could still serve as a valuable predictor in
the model. The moderate positive correlation between log-transformed
road length within a 1 km radius and PM2.5 values (0.2053322) points
towards road length being a good indicator of vehicular emissions, which
are a major source of air pollution. The correlation between the
Satellite Data (optical_depth) and the PM2.5 values is 0.3498.This
represents a moderate positive correlation, suggesting a more
substantial relationship where higher optical depth readings are
associated with higher PM2.5 values.The correlation betwee the Chemical
Modeling Data (CMAQ) and the PM2.5 values is 0.4662. This is the
strongest correlation observed among the predictors. It indicates a
moderate to strong positive correlation, suggesting that higher CMAQ
values are significantly associated with higher PM2.5 values.

```{r}
#linear regression 
model_popdens <- lm(value ~ popdens_zcta, data = pm)
model_urbanization <- lm(value ~urc2013, data = pm)
model_roaddens <- lm(value ~ log_RoadLength_1kmRadius, data = pm)
model_optdep <- lm(value ~ optical_depth, data = pm)
model_CMAQ <- lm(value ~ CMAQ, data = pm)

summary(model_CMAQ)

#Visualizations 
# Model with population density
ggplot(pm, aes(x = popdens_zcta, y = value)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Model of Value vs Population Density",
       x = "Population Density (ZCTA)",
       y = "Value") +
  theme_minimal()

# Model with urbanization
ggplot(pm, aes(x = urc2013, y = value)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Model of Value vs Urbanization",
       x = "Urbanization (URC 2013)",
       y = "Value") +
  theme_minimal()

# Model with road density
ggplot(pm, aes(x = log_RoadLength_1kmRadius, y = value)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Model of Value vs Log Road Length within 1km Radius",
       x = "Log Road Length within 1km Radius",
       y = "Value") +
  theme_minimal()

# Model with optical depth
ggplot(pm, aes(x = optical_depth, y = value)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Model of Value vs Optical Depth",
       x = "Optical Depth",
       y = "Value") +
  theme_minimal()

# Model with CMAQ
ggplot(pm, aes(x = CMAQ, y = value)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Model of Value vs CMAQ",
       x = "CMAQ Value",
       y = "Value") +
  theme_minimal()

```

Here we plotted the linear regression graph of each of the 5 factors
against pm 2.5 value.

**Result**: We can see that there is a slight positive relationship
between pm 2.5 value and population density, road density (represented
by log raod lenth within 1 km radius), and optic depth. There is a
pretty significant positive relationship between pm 2.5 value and CMAQ
value, and a slight negative realtionship between pm 2.5 value and
urbanization, which is a suprising result.

```{r}
# Model Evaluation
summary(model_popdens)$coefficients
summary(model_urbanization)$coefficients
summary(model_roaddens)$coefficients
summary(model_optdep)$coefficients
summary(model_CMAQ)$coefficients

# Extract model summaries using glance
summary_popdens <- glance(model_popdens)
summary_urbanization <- glance(model_urbanization)
summary_roaddens <- glance(model_roaddens)
summary_optdep<- glance(model_optdep)
summary_CMAQ <- glance(model_CMAQ)

# Add a column to each summary to indicate the model
summary_popdens$model <- "Population Density"
summary_urbanization$model <- "Urbanization Level"
summary_roaddens$model <- "Road Density"
summary_optdep$model <- "Satellite Data"
summary_CMAQ$model <- "Chemical Modeling Data"

factors <- c("PopDens", "Urbanization", "RoadDens", "OptDep", "CMAQ")

# Bind the rows and add the 'factor' column
combined_summary <- bind_rows(
  summary_popdens %>% mutate(factor = factors[1]),
  summary_urbanization %>% mutate(factor = factors[2]),
  summary_roaddens %>% mutate(factor = factors[3]),
  summary_optdep %>% mutate(factor = factors[4]),
  summary_CMAQ %>% mutate(factor = factors[5])
)

combined_summary <- combined_summary %>% 
  select(factor, everything())

combined_summary
```

To better interpret the result from our linear regression model above,
we printed the result table of our linear regression models. The
Pr(\>\|t\|) column here can tell us the probability that the result is
significant.

**Result:**

Coefficient for Population Density: 1.246e-04 This model suggests a
slight positive relationship between population density and the response
variable. However, the relatively small coefficient indicates that
population density is not a strong predictor for estimating the air
pollution. The R-squared value of 0.0177 reflects this, showing that
only a small percentage of the variability in the response variable is
explained by population density alone.

Coefficient for Urbanization Level: -0.4266 The negative coefficient
here implies that as the urbanization level increases, the value
decreases. This could suggest that more urbanized areas are associated
with lower values of the response variable. The R-squared value of
0.0629, though higher than that for population density, still accounts
for only a modest amount of the variance, which indicates that
urbanization, while a significant factor, is not the sole determinant of
air pollution levels.

Coefficient for Road Density: 0.6716 This positive coefficient indicates
that an increase in road density is associated with an increase in the
response variable. The magnitude of the coefficient suggests a moderate
relationship, indicating that road density is a good predictor for
building the air pollution forecasting model. The model's R-squared
value of 0.0422 supports this, showing that road density explains a
larger proportion of the variance than population density but is still a
relatively modest predictor.

Coefficient for Optical Depth: 0.0462 A positive coefficient here
suggests that higher optical depth readings correlate with higher values
of the response variable. This relationship is more pronounced compared
to the previous models, as indicated by the coefficient's magnitude. An
R-squared value of 0.1224 indicates that optical depth is a significant
predictor, explaining a more considerable portion of the variance in the
response variable.

Coefficient for CMAQ: 0.4053 This model shows a significant positive
relationship between CMAQ values and the response variable, implying
that higher CMAQ values are associated with higher values of the
response variable. The R-squared value of 0.2173 is quite significant,
indicating that this model has a strong predictive power and that CMAQ
is a robust predictor of air pollution levels.

```{r}
# Split the dataset into training data and testing data
set.seed(1234) #ensures we all get the exact same random split
pm_split <- rsample::initial_split(data = pm, prop = 2/3)
pm_split
train_pm <- rsample::training(pm_split)
test_pm <- rsample::testing(pm_split)

count(train_pm, state)
```

```{r}
# Training the random forest model
model_rf <- randomForest(value ~ popdens_zcta + 
                        urc2013 + 
                        log_RoadLength_1kmRadius + 
                        optical_depth + 
                        CMAQ, 
                      data = train_pm)

# Evaluating the model
predictions_rf <- predict(model_rf, test_pm)
mae_rf <- mean(abs(predictions_rf - test_pm$value))
rmse_rf <- sqrt(mean((predictions_rf - test_pm$value)^2))

print(paste("MAE_RF:", mae_rf))
print(paste("RMSE_RF:", rmse_rf))
```

```{r}
# Training the Linear Regression Model 
model_lm <- lm(value ~ popdens_zcta + 
                 urc2013 + 
                 log_RoadLength_1kmRadius + 
                 optical_depth + 
                 CMAQ, 
               data = train_pm)

# Evaluating the model
predictions_lm <- predict(model_lm, test_pm)
mae_lm <- mean(abs(predictions_lm - test_pm$value))
rmse_lm <- sqrt(mean((predictions_lm - test_pm$value)^2))

print(paste("MAE_LM:", mae_lm))
print(paste("RMSE_LM:", rmse_lm))

```

In order to find a model that can have better accuracy. We used a
classic ranfom forest model, trained and testes its performance on this
dataset. Here we splited the Train dataset to be 2/3 of the dataset (584
observations), and Test dataset is 1/3 of the dataset (292 observations)
Total (pm): 584+292 = 876 observations.

**Result:** The MAE for the Linear Model is 1.5549, which means that, on
average, the absolute difference between the predicted and actual values
is about 1.5549 units. When compared to the Random Forest model (MAE of
1.4558), the Linear Model has a slightly higher MAE. This suggests that
the Linear Model, on average, has marginally larger errors in its
predictions.The RMSE for the Linear Model is 2.1797, indicating that the
standard deviation of the prediction errors is about 2.1797 units.
Comparatively, the Random Forest model had an RMSE of 1.9818, which is
lower than the RMSE of the Linear Model. This indicates that the Linear
Model has larger prediction errors and is less accurate than the Random
Forest model, especially when it comes to predicting values with larger
deviations from the mean.

In conclusion, our Random Forest model uses data about population
density, urbanization, road density, satellite pollution data, and
chemical modeling data to predict US annual average air pollution
concentrations at the granularity of zip code regional levels with great
accuracy with a MAE of 1.456.

### Extended Data Analysis

Refining our principal analytical question, we now concentrate on the
relationship between air quality and mortality due to respiratory
diseases at the state level. This pivot is necessitated by the
granularity of our newly sourced dataset from the Centers for Disease
Control and Prevention (CDC), which provides insights into weekly counts
of deaths by state for select causes, including respiratory diseases.

Our revised hypothesis suggests that states burdened with higher air
pollutant concentrations will report higher mortality rates from
respiratory conditions. To test this hypothesis, we will implement
statistical tests such as t-tests or z-tests. These will compare the
mean mortality rates across states categorized by varying levels of
pollution. This statistical approach will elucidate the significance of
the differences in mortality rates and allow us to assess the public
health implications of air pollution across the United States.

The CDC dataset is accessible here: [Weekly Counts of Deaths by State
and Select
Causes](https://data.cdc.gov/NCHS/Weekly-Counts-of-Deaths-by-State-and-Select-Causes/3yf8-kanr/about_data).

```{r}
# Data Cleaning for health dataset
health <- readr::read_csv(here::here("OCS_data", "data","raw", "health_data.csv"))
health <- health %>%
  select(`Chronic lower respiratory diseases (J40-J47)`, 
         `Other diseases of respiratory system (J00-J06,J30-J39,J67,J70-J98)`, 
         `Jurisdiction of Occurrence`) |>
  rename(
    CLow = `Chronic lower respiratory diseases (J40-J47)`,
    Other = `Other diseases of respiratory system (J00-J06,J30-J39,J67,J70-J98)`,
    State = `Jurisdiction of Occurrence`
  )|>
  filter(State != "United States")|>
  mutate(across(everything(), ~replace_na(., 0)))|>
  group_by(State) |>
  summarise(Other = sum(Other),CLow = sum(CLow))

pm <- readr::read_csv(here::here("OCS_data", "data","raw", "pm25_data.csv"))
pm_ext <- pm %>%
  select(state, value) |>
  rename(State = `state`)|>
  group_by(State) |>
  summarise(mean_value = mean(value, na.rm = TRUE))

ext <- left_join(pm_ext, health, by = "State")
ext <- drop_na(ext)
ext
```

To clean the data for our extended analysis, I imported our health data,
and extracted columns pertinent to chronic lower respiratory diseases,
other respiratory conditions, and the respective locations. I also
eliminated entries that aggregated data at the national level, ensuring
our focus remained state-specific. After addressing missing values by
substituting them with zeroes, I aggregated disease counts by state.
Similarly, I processed our PM2.5 dataset to calculate the average
concentration per state. Merging these datasets on the state attribute
resulted in a consolidated dataset, 'ext', stripped of any missing
values, thereby laying a robust foundation for subsequent correlation
analysis and t-tests.

```{r}
high_pollution <- ext %>%
  mutate(high_pollution = mean_value > median(mean_value))
t_test_result1 <- t.test(CLow ~ high_pollution, data = high_pollution)
t_test_result2 <- t.test(Other ~ high_pollution, data = high_pollution)
print(t_test_result1)
print(t_test_result2)
```

Here I first binarily labeled all of the States according to how poluted
it is compared to national average. Then I ran a Two Sample t-tests on
the relationship between pm2.5 value and both chronic respiratory
diseases and other repiratory diseases.

**Result:** The results of the Two Sample t-tests indicate that there is
a statistically significant difference in the mean number of deaths from
both 'Chronic lower respiratory diseases' (CLow) and 'Other diseases of
respiratory system' between states classified as having high and low
pollution levels.

For CLow, the p-value is 0.005602, which is less than the conventional
alpha level of 0.05, suggesting a significant difference in means.
Similarly, for the Other category, the p-value is 0.001536, also
indicating a significant difference in means.

The negative t-values in both tests suggest that the group labeled TRUE
(states with higher pollution) has higher mean deaths compared to states
with lower pollution. The confidence intervals do not contain zero,
further supporting the conclusion that the difference in means is likely
not due to random chance.

```{r}
lm_CLow <- lm(CLow ~ mean_value, data = ext)
summary(lm_CLow)

lm_Other <- lm(Other ~ mean_value, data = ext)
summary(lm_Other)
```

```{r}
ggplot(ext, aes(x = mean_value, y = CLow)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Model of PM2.5 vs Death by Chronic Respiratory Disease",
       x = "Mean PM 2.5 Value by State",
       y = "Death by Chronic Respiratory Disease") +
  theme_minimal()

ggplot(ext, aes(x = mean_value, y = Other)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Model of PM2.5 vs Death by Other Respiratory Disease",
       x = "Mean PM 2.5 Value by State",
       y = "Death by Other Respiratory Disease") +
  theme_minimal()
```

To further explore the relationship between PM 2.5 levels and the number
of deaths from respiratory diseases, we conducted linear regression
analyses.

**Result:** The results indicate a significant positive relationship;
for every unit increase in PM 2.5, there is an estimated increase of
3,463 deaths from chronic lower respiratory diseases and 1,107 deaths
from other respiratory diseases. The R-squared values---0.1625 for CLow
and 0.2156 for Other---suggest that PM 2.5 levels explain a moderate
proportion of the variance in mortality rates. Interestingly, the
negative intercepts imply that the model predicts a negative number of
deaths at zero PM 2.5 level, highlighting the limitation of using linear
models for such data and the potential influence of other unaccounted
factors. Nonetheless, the significant coefficients reaffirm the
detrimental impact of increased PM 2.5 levels on respiratory health.

Therefore, these results support the hypothesis that higher air
pollution levels are associated with increased mortality from all kinds
of respiratory diseases at the state level.

## Conclusion

In our study, we explore the data about the average air pollution
concentrations across the United States at the granularity of zip code
regions. The study shows that there is a correlation between PM2.5
(respirable particulate matter) concentrations in different regions of
the United States and factors such as population density, level of
urbanization, road density, optical depth, and CMAQ model values.
Through a linear regression analysis using the R language, we can see
that After our prediction we found a moderate positive correlation
between road density and PM2.5 values, implying that areas with higher
road density may have more air pollution due to vehicle emissions.
Meanwhile, there is a strong positive correlation between the CMAQ model
values and PM2.5 values, indicating that the CMAQ model is a powerful
predictive tool that can accurately predict PM2.5 pollution levels. In
terms of model evaluation, the Random Forest model performed slightly
better than the linear regression model with lower MAE and RMSE values,
indicating that it has better predictive accuracy and generalization
ability when dealing with this type of data. Although the factors had
different levels of influence on PM2.5 pollution levels, the CMAQ model
and road density were two important predictors.

Our extended analysis focused on the hypothesis of the study was that
states with higher concentrations of air pollutants would report higher
mortality rates from respiratory diseases. To test this hypothesis, we
implemented t-test and linear regression model to compare average
mortality rates between states with pollution levels. These statistical
methods will shed light on differences in mortality rates and allow us
to assess the public health impacts of air pollution across the United
States. The results of the study show that there are statistically
significant differences, with higher mortality rates due to chronic
lower respiratory disease and other respiratory illnesses in states with
higher concentrations of pm 2.5.

### References

1.  Frontiers \| Environmental and Health Impacts of Air Pollution: A
    Review.
    <https://www.frontiersin.org/articles/10.3389/fpubh.2020.00014/full.Final>
    Report

2.  Yanosky, J. D. et al. Spatio-temporal modeling of particulate air
    pollution in the conterminous United States using geographic and
    meteorological predictors. Environ. Health 13, 63 (2014).

3.  Ellis, P. S. COGS 137 - 15-cs02-data.
    <https://cogs137.github.io/website/content/lectures/15-cs02-data.html#pm-2.5-data>
    (2023).

4.  Weekly Counts of Deaths by State and Select Causes, 2014-2019 \|
    Data \| Centers for Disease Control and Prevention.
    <https://data.cdc.gov/NCHS/Weekly-Counts-of-Deaths-by-State-and-Select-Causes/3yf8-kanr/about_data>.
